---
title: "Cheese!"
description: |
  Analysis of Cheese data from TidyTuesday
author: Jack Susank
date: February 15, 2025
format: html
echo: FALSE
execute:
  warning: false
  message: false
---

This is an analysis of Cheese data sourced from TidyTuesday's June 4th, 2024 data release. The plot compares the fat and calcium contents different types of cheese. To download the data, visit [this github repo](https://github.com/rfordatascience/tidytuesday/blob/main/data/2024/2024-06-04/readme.md).
This dataset was compiled using data from [Cheese.com](https://www.cheese.com/) and its creation was inspired by the [polite package](https://dmi3kno.github.io/polite/). 

```{r setup-pat, include=FALSE}
readRenviron("~/.Renviron")
```


```{r}
cheeses <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-06-04/cheeses.csv')
```

```{r}
library(ggplot2)
library(stringr)
library(dplyr)


cleaned_cheeses <- cheeses |>
  select(fat_content, calcium_content, cheese) |>
  mutate(calcium_content = as.integer(str_replace(calcium_content, " mg/100g", ""))) |>
  mutate(fat_content = as.integer(str_replace(fat_content, "%", ""))) |>
  filter(!is.na(fat_content) & !is.na(calcium_content) & !is.na(cheese))
```


```{r, alt-text = Scatterplot showing fat content versus calcium content for various cheeses, with each cheese type represented by a different color and a blue linear trend line indicating an overall weak negative relationship. Cheeses listed include Limburger, Petida, Anthotyro, and Provoleta.}

# Scatterplot of fat content vs. calcium content for various cheeses
ggplot(cleaned_cheeses, aes(x = fat_content, y = calcium_content, color = cheese)) +
  # Plot individual cheese data points, colored by cheese type
  geom_point() +
  
  # Add a linear trend line (without a confidence interval)
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(
    title = "Fat versus Calcium content of Various Cheeses",
    x = "Fat Content (%)",
    y = "Calcium Content (mg/100g)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "gray20"),
    panel.grid.major = element_line(color = "gray85"),
    panel.grid.minor = element_blank()
  )
```


This plot visualizes the relationship between fat and calcium content across various types of cheese. Each point represents a specific cheese, and a linear trend line suggests a weak negative correlation.