---
title: "Water Insecurity"
description: |
  Analysis of Water Insecurity data from TidyTuesday
author: Jack Susank
date: February 15, 2025
format: html
echo: FALSE
execute:
  warning: false
  message: false
---

This is an analysis of Water Insecurity data sourced from TidyTuesday's January 28th, 2025 data release. The plot depicts the water insecurity of US Counties in 2022. To download the data, visit [this github repo](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-01-28/readme.md).
This dataset was compiled by [Niha Pereira](https://github.com/nnpereira) using the [tidycensus](https://CRAN.R-project.org/package=tidycensus) package for R and [this blog walkthrough](https://waterdata.usgs.gov/blog/acs-maps/). For more information on how to access census data from sources such as the U.S. Census Bureau, the Decennial Census, the American Community Survey (ACS), and the Household Pulse Survey, see the walkthrough. 


```{r}
library(ggplot2)

tuesdata <- tidytuesdayR::tt_load(2025, week = 4)

water_insecurity_2022 <- tuesdata$water_insecurity_2022

water_insecurity_2022 <- na.omit(water_insecurity_2022)

```

```{r}
ggplot(water_insecurity_2022, aes(x = total_pop, y = percent_lacking_plumbing)) +
  geom_point(color = "blue", size = 2, alpha = 0.5) +
  labs(
    title = "Water Insecurity in US Counties",
    x = "Total Population",
    y = "Percent Lacking Plumbing"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "gray20"),
    panel.grid.major = element_line(color = "gray85"),
    panel.grid.minor = element_blank()
  ) +
  scale_x_continuous(labels = scales::comma)

```
